Diffusion Process and its optimization

### 2015-04-21 opt-methods.tex
Primeros cambios, Maribel

\title{Using evolutionary algorithms to optimize the estimation of a diffusion process}
%
% You need the command \numberofauthors to handle the 'placement
% and alignment' of the authors beneath the title.
%
% For aesthetic reasons, we recommend 'three authors at a time'
% i.e. three 'name/affiliation blocks' be placed beneath the title.
%
% NOTE: You are NOT restricted in how many 'rows' of
% "name/affiliations" may appear. We just ask that you restrict
% the number of 'columns' to three.
%
% Because of the available 'opening page real-estate'
% we ask you to refrain from putting more than six authors
% (two rows with three columns) beneath the article title.
% More than six makes the first-page appear very cluttered indeed.
%
% Use the \alignauthor commands to handle the names
% and affiliations for an 'aesthetic maximum' of six authors.
% Add names, affiliations, addresses for
% the seventh etc. author(s) as the argument for the
% \additionalauthors command.
% These 'additional authors' will be output/set for you
% without further effort on your part as the last section in
% the body of your article BEFORE References or any Appendices.

@@

\begin{abstract}
Diffusion processes are stochastic models, % of what? -- JJ
 usually employed for
modeling and forecasting in observed paths that vary along time under
random environment. % Don't understand -- clarify - JJ
 Such models are useful if we are able to know the values of the
 parameters involved. The maximum likelihood estimation of the
 parameters requires a system of equations that, for some cases, has
 no explicit solution; thus, the solution needs be estimated. We will
 use optimization methods to find those solutions: an Iterative
 Method, an algorithm based on Newton-Raphson solver, a Variable
 Neighbourhood Search method, a Simulated Annealing algorithm and an
 Evolutionary Algorithm. We generate four data sets following a
 Gompertz-lognormal diffusion process using different noise
 levels. % are these data sets complete? They correspond to what? - JJ
The methods are applied to these data sets in order to
 estimate the parameters that rule the diffusion process. Results show
 that bio-inspired methods yield suitable solutions for the problem in
 all four data sets, even when the noise level increases. %noise is
                                %first mentioned here. Why is it
                                %important? - JJ
 On the other hand,
 some analytical methods as Newton-Raphson or the Iterative Method do
 not always solve the problem if their scores depend on the
 starting point for initial solution or the noise level hinders the
 resolution of the problem. In these cases, the bio-inspired
 algorithms remain  a suitable and 
reliable approach.

@@

Many probabilistic models are frequently used for modelling natural
growth-patterns and making forecasts % of what? - JJ
. Although there exists a lot of
different growth-patterns, and many different ways of modeling them,
among the most widely
studied and useful models are diffusion processes. % cita - JJ
 They are frequently
employed in biological growth modelling. \cite{lognormal}.

Diffusion process is an usual mathematical formulation of a model for population growth in a randomly
varying environment \cite{lognormal}. %again same citation?  - JJ
 These models also describe how a population adopts a new innovation, technology or behaviour \cite{Myrskyla2013}. With these kind of models, we have the transition density function (t.d.f.) which express the chance of observing
any size for the population at any time when the initial size is specified. With the t.d.f we also are able to determine the time at that population first reaches a fixed size. Moreover, we can forecast future values of the population size as well as confidence intervals for the predictions.

In this paper we focus our attention on the Gompertz-lognormal diffusion process. It can be seen as a mixture between a Gompertz-type process \cite{Gut07b} and a lognormal process \cite{GutCy}. The first one models precesses with trends showing enclosed growth patterns, as biological growth with finite space or resources. The second one is useful for modelling not bounded growth with exponential trend.

@@

This point take us to propose different approaches to find the estimates of the parameters, maximising the likelihood function via numerical approximations or finding the estimates with other strategy. In Section \ref{sec:soa} we review the state of art about optimization methods. We describe the proposed methods in Section \ref{sec:methods}. We compare afterwards, at Section \ref{sec:pathsStudy} all the methods using four simulated data set with different noise levels. Finally, in Section \ref{sec:Conclusion} we establish the main conclusions of the study.

@@

\begin{enumerate}
\item First, we fix an initial value for parameter $\beta$ and the method uses it to get an initial set of solutions ($\{\hat{m}; \hat{\beta}; \hat{c}; \hat{\sigma}^2\}$) through assessing (\ref{estima}).
\item Second, the method calculates $\beta'$ using the point of the path where the second inflection point is set, for improving the $\beta$ value.
\item Third,  the method goes to step 1 and sets $\beta'$ as $\beta$. The process iterates until $|\beta-\beta'|<10^{-8}$.

@@

We have selected, as stopping criteria, a maximum tolerance level in the iterative process. In particular we consider a threshold of $10^{-8}$.

@@

In this case, the local search (2.b.ii) is the Neldel-Mead algorithm which does not carry out the gradient computation. The understanding for this choice is that if function (\ref{funcion}) is not derivable, the gradient does not exist. Moreover, we repeat second step for each range, $10$ times as maximum to assure the convergence to the optimum solution inside the current range. The stopping rule equals maximum tolerance to $10^{-8}$.

@@

\begin{enumerate}
\item At the beginning it generates a random solution as starting point. The algorithm generates a new solution from the current one by means of a change operator (new state generator).
\item If the new solution is better, it is accepted. However, if it is worse, the algorithm accepts it using the probability of $p_a = e^{ - \Delta f / T }$ , where $\Delta f$  is the increment of the cost function and $T$ is the current value of temperature parameter. The temperature is then decreased using the temperature cold function $f_T$. The decrement rule used for this paper is $T_{n+1} = \alpha T_n$ , where $\alpha$ is equal to $0.95$ proposed by Kirkpatrick et al. \cite{Kirkpatrick83}.  For each temperature value, the algorithm test $32$ different solutions.
\item The algorithm starts with temperature equal to $10$ and ends when the temperature decreases according to the $f_T$ until $0.0001$.
\end{enumerate}

>>>>

\title{Diffusion Process and its optimization}
%
% You need the command \numberofauthors to handle the 'placement
% and alignment' of the authors beneath the title.
%
% For aesthetic reasons, we recommend 'three authors at a time'
% i.e. three 'name/affiliation blocks' be placed beneath the title.
%
% NOTE: You are NOT restricted in how many 'rows' of
% "name/affiliations" may appear. We just ask that you restrict
% the number of 'columns' to three.
%
% Because of the available 'opening page real-estate'
% we ask you to refrain from putting more than six authors
% (two rows with three columns) beneath the article title.
% More than six makes the first-page appear very cluttered indeed.
%
% Use the \alignauthor commands to handle the names
% and affiliations for an 'aesthetic maximum' of six authors.
% Add names, affiliations, addresses for
% the seventh etc. author(s) as the argument for the
% \additionalauthors command.
% These 'additional authors' will be output/set for you
% without further effort on your part as the last section in
% the body of your article BEFORE References or any Appendices.

@@

\begin{abstract}
Diffusion processes are stochastic models, % of what? -- JJ
 usually employed for
modeling and forecasting in observed paths that vary along time under
random environment. % Don't understand -- clarify - JJ
 Such models are useful if we are able to know the values of the
 parameters involved. The maximum likelihood estimation of the
 parameters requires a system of equations that, in some cases, has
 no explicit solution; thus, the solution must be estimated. In this paper, we propose several ways to solve this problem using some optimization methods to find those solutions: an Iterative
 Method, an algorithm based on Newton-Raphson solver, a Variable
 Neighbourhood Search method, a Simulated Annealing algorithm and an
 Evolutionary Algorithm. Four data sets following a
 Gompertz-lognormal diffusion process have been tested with different noise
 levels. % are these data sets complete? They correspond to what? - JJ
The methods are applied to these data sets in order to
 estimate the parameters that rule the diffusion process. Results show
 that bio-inspired methods yield suitable solutions for the problem in
 all four data sets, even on instances of the problem with higher noise level. 
 %noise is
 %first mentioned here. Why is it
 %important? - JJ
 Moreover, some analytical methods as Newton-Raphson or the Iterative Method do
 not always find the solution for the problem because their scores depend on the
 starting point for initial solution or because the noise level hinders the
 resolution of the problem. In these cases, the bio-inspired
 algorithms remain a suitable and reliable approach.

@@

Many probabilistic models are frequently used for modelling natural
growth-patterns and making forecasts of data.% of what? - JJ
Although there exists a lot of
different growth-patterns, and many different ways of modeling them,
among the most widely
studied and useful models are diffusion processes \cite{rogerswilliams1} % cita - JJ
usually employed in biological growth modelling \cite{Kaandorp:400332}.

Diffusion process is an usual mathematical formulation of a model for population growth in a randomly
varying environment \cite{lognormal}. %again same citation?  - JJ
 These models can also be used for describing how a population adopts a new innovation, technology or behaviour \cite{Myrskyla2013}. These kind of models set the transition density function (t.d.f.) which express the chance of observing
any size for the population at any time when the initial size is specified. The t.d.f also determines the time at that population first reaches a fixed size. Moreover, we can forecast future values of the population size as well as confidence intervals for the predictions.

In this paper we focus our attention on the Gompertz-lognormal \cite{eurocast2013} diffusion process because it it has not been studied enough and it arises as base for several important diseases spreading. The details of the process are detailed in section \ref{sec:gompertz}.

The rest of the paper is organised as follow: Section \ref{sec:soa} reviews the state of art about optimization methods. Section \label{sec:gompertz}  introduces the Gompertz-lognormal process. The proposed methods are detailed in Section \ref{sec:methods}. Afterwards, at Section \ref{sec:pathsStudy} the experiments, the results using four simulated data sets with different noise levels are included. Finally, in Section \ref{sec:Conclusion} we establish the main conclusions of the study.

\section{Gompertz-Lognormal Difussion Process}
\label{sec:gompertz}

Gompertz-lognormal diffusion process is childlike comparing it with other classical processes as [FALTAN LOS CLÁSICOS Y REFERENCIA A ELLOS]. 
It can be seen as a mixture between a Gompertz-type process \cite{Gut07b} and a lognormal process \cite{GutCy}. The first one models processes with trends showing enclosed growth patterns, as biological growth with finite space or resources. The second one is useful for modelling not bounded growth with exponential trend.

@@

This point take us to propose different approaches to find the estimates of the parameters, maximising the likelihood function via numerical approximations or finding the estimates with other strategy.

@@

\begin{enumerate}
\item First, we fix an initial value for parameter $\beta$ and the method uses it to get an initial set of solutions ($\{\hat{m}; \hat{\beta}; \hat{c}; \hat{\sigma}^2\}$) through assessing (\ref{estima}).
\item Second, the method calculates $\beta'$ using the point of the path where the second inflection point is set, for improving the $\beta$ value.
\item Third,  the method goes to step 1 and sets $\beta'$ as $\beta$. The process iterates until $|\beta-\beta'|<10^{-8}$.[POR QUE ESTE VALOR]

@@

We have selected, as stopping criteria, a maximum tolerance level in the iterative process. In particular we consider a threshold of $10^{-8}$. [PORQUÉ ESTE VALOR]

@@

In this case, the local search (2.b.ii) is the Neldel-Mead algorithm \cite{NelderMead65} which does not carry out the gradient computation. The understanding for this choice is that if function (\ref{funcion}) is not derivable, the gradient does not exist. Moreover, we repeat second step for each range, $10$ times as maximum to assure the convergence to the optimum solution inside the current range. The stopping rule equals maximum tolerance to $10^{-8}$.

@@

\begin{enumerate}
\item At the beginning it generates a random solution as starting point. The algorithm generates a new solution from the current one by means of a change operator (new state generator).
\item If the new solution is better, it is accepted. However, if it is worse, it is accepted with a probability of $p_a = e^{ - \Delta f / T }$ , where $\Delta f$  is the increment of the cost function and $T$ is the current value of temperature parameter. The temperature is then decreased using the temperature cold function $f_T$. The decrement rule used for this paper is $T_{n+1} = \alpha T_n$ , where $\alpha$ is equal to $0.95$ proposed by Kirkpatrick et al. \cite{Kirkpatrick83}.  For each temperature value, the algorithm test $32$ different solutions.
\item The algorithm starts with temperature equal to $10$ and ends when the temperature decreases according to the $f_T$ until $0.0001$.
\end{enumerate}
